cardinalR: Generating interesting high-dimensional data structures

### 2025-05-07 paper-cardinalR.tex
Try to fix the issues of compiling the Latex code

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in 2D, with optional noise dimensions to embed the structure in a \$p\$-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch \(i\) is constructed using:

\(X\_1 \sim \mathcal{U}(-2, 2)\)
\(X\_2 = \exp(\pm s\_i \cdot X\_1) + \epsilon\),

where:
\(\epsilon \sim \mathcal{U}(0, 0.1)\) adds local jitter,

\(s\_i \in [0.5, 2]\) is randomly sampled for each branch.

The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected about \(x\)-axis) and even-numbered branches: growth pattern.

For \(p > 2\), Gaussian noise is added to embed the \(2\text{-}D\) branches in higher dimensions:

\[
X\_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
\]

These represent irrelevant or noisy dimensions.

\begin{verbatim}
expbranches <- gen_expbranches(n = 1000, p = 4, k = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-expbranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `expbranches` data.}\label{fig:fig-expbranches-proj}
\end{figure}

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in a \(p\)-dimensional space. The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from \(X\_1 \sim \mathcal{U}(-2, 8)\$, \$X\_2 = 0.5 \cdot X\_1 + \epsilon\). The Branch 2 is generated from \(X\_1 \sim \mathcal{U}(-6, 2)\$, \$X\_2 = -0.5 \cdot X\_1 + \epsilon\). \(\epsilon \sim \mathcal{U}(0, 0.5)\).

Branches 3 to \(k\)** are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

\(X\_1\) values span a short segment (\(x\_{start}\) to \(x\_{start} + 1\)),

\(X\_2 = s\_i \cdot (X\_1 - x\_{start}) + y\_{start} + \epsilon\) where \(s\_i\) is a sampled slope from a filtered set and \(\epsilon \sim \mathcal{U}(0, 0.2)\).

This setup yields \(k\) spatially distinct, noisy linear trajectories.

For \(p > 2\), the remaining dimensions are pure noise:

\[
X\_j \sim \mathcal{N}(0, 0.05^2)\$ for \$j = 3, \dots, p.
\]

\begin{verbatim}
linearbranches <- gen_linearbranches(n = 1000, p = 4, k = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-linearbranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `linearbranches` data.}\label{fig:fig-linearbranches-proj}
\end{figure}

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in a \(p\)-dimensional space. The underlying geometry lies in the first two dimensions, while the remaining \(p - 2\) dimensions contain Gaussian noise.

Branch 1 is generated from \(X\_1 \sim \mathcal{U}(0, 1)\), and \(X\_2 = 0.1 \cdot X\_1 + 1 \cdot X\_1^2 + \epsilon\), where \(\epsilon \sim \mathcal{U}(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from \(X\_1 \sim \mathcal{U}(-1, 0)\), and \(X\_2 = 0.1 \cdot X\_1 - 2 \cdot X\_1^2 + \epsilon\), where \(\epsilon \sim \mathcal{U}(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane.

Branches \(3\) to \(k\)** are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: \(X\_1 \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X\_1\):\(X\_1 \in [x\_{\text{start}}, x\_{\text{start}} + 1]\), and has the structure:

\[
X_2 = 0.1 \cdot X_1 - s_i \cdot (X_1^2 - x_{\text{start}}) + y_{\text{start}}, \quad s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}.
\]

Here, \(s\_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x\_{\text{start}}, y\_{\text{start}})\) is the starting point from the existing structure.

This construction yields \(k\) spatially connected, nonlinear branches with varying curvature.

For \(p > 2\), the remaining dimensions are pure Gaussian noise:

\[
X_j \sim \mathcal{N}(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
\]

\begin{verbatim}
curvybranches <- gen_curvybranches(n = 1000, p = 4, k = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-curvybranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `curvybranches` data.}\label{fig:fig-curvybranches-proj}
\end{figure}

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in a \(p\)-dimensional space. Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((x\_{i1}, x\_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n\_i\) points are generated, where \(\sum\_{i=1}^k n\_i = n\). The structure follows:

\[
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5),
\]

where \(s\_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s\_i = 1\). When sampling with replacement, \(s\_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\).

For \(p > 2\), the remaining dimensions contain independent Gaussian noise:

\[
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

\begin{verbatim}
orglinearbranches <- gen_orglinearbranches(n = 1000, p = 4, k = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-orglinearbranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `orglinearbranches` data.}\label{fig:fig-orglinearbranches-proj}
\end{figure}

The \texttt{gen\_orgcurvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in a \(p\)-dimensional space. Each branch is constructed in a unique or repeated \(2\text{-}D\) subspace of the \(p\)-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining \(p - 2\) dimensions contain Gaussian noise.

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((x\_{i1}, x\_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s\_i\) is used to vary the curvature:

\begin{itemize}
\tightlist
\item
  If \(k \leq \binom{p}{2}\): \$s\_i = 1 \$
\item
  If \(k > \binom{p}{2}\): \(s\_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\)
\end{itemize}

Each branch contains \(n\_i\) points such that \(\sum\_{i=1}^k n\_i = n\), where the vector \((n\_1, \dots, n\_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\).

Within its assigned subspace \((x\_{i1}, x\_{i2})\), branch \(i\) is defined by:

\[
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5)
\]

This forms a smooth downward-opening parabola in the plane defined by \((x\_{i1}, x\_{i2})\), with the degree of curvature controlled by \(s\_i\).

For \(p > 2\), all remaining dimensions are independent Gaussian noise:

\[
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

\begin{verbatim}
orgcurvybranches <- gen_orgcurvybranches(n = 1000, p = 4, k = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-orgcurvybranches-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `orgcurvybranches` data.}\label{fig:fig-orgcurvybranches-proj}
\end{figure}

@@

To simulate a cone-shaped structure in arbitrary dimensions, we define a function \texttt{gen\_cone(n,\ p,\ h,\ ratio)}, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates \texttt{n} points in a \texttt{p}-dimensional space, where the last dimension (\texttt{X\_p}) represents height along the cone's axis, and the remaining dimensions define a shrinking hyperspherical cross-section as one moves toward the tip.

Points along the height axis are drawn from an exponential distribution to increase density near the tip:

\[
X_p = h_i \sim \text{Exp}(\lambda = 2/h), \quad \text{truncated at } h
\]

The effective radius at height \(h_i\) decreases linearly from base to tip, controlled by a shape ratio parameter:

\[
r_i = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) \cdot \frac{h_i}{h}, \quad \text{where } r_{\text{min}} = \text{ratio},\ r_{\text{max}} = 1
\]

For each point, a direction is sampled uniformly from a \((p-1)\)-dimensional hypersphere using generalized spherical coordinates (angles). The radial coordinates are scaled by \(r_i\), which ensures a conical taper. Specifically, when \(p = 3\), this scaling results in a classic \(3\text{-}D\) cone. However, for \(p > 3\), the introduction of additional angular components allows for a smooth extension into higher dimensions, preserving the conical shape while accommodating the complexities of multi-dimensional geometry.

@@

The \texttt{gen\_gridcube(n,\ p)} constructs a regular grid using approximately \(n\) points by evenly partitioning each of the \(p\) dimensions. Specifically, the number of grid points along each axis is determined by \texttt{gen\_nproduct(n,\ p)}, which calculates the nearest balanced integer factors whose product is close to \(n\). The full grid is generated via \texttt{expand\_grid()}, giving Cartesian combinations of these evenly spaced positions.

Each variable \(X\_j\) represents integer grid coordinates (e.g., 1, 2, 3, \ldots) in dimension \(j\):

\[
X\_j \in {1, \dots, k\_j},\text{ where }\prod\_{j=1}^p k\_j \approx n.
\]

This results in a hypercube-shaped lattice in \(p\)-dimensional space.

\begin{verbatim}
gridcube <- gen_gridcube(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-gridcube-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gridcube` data.}\label{fig:fig-gridcube-proj}
\end{figure}

The \texttt{gen\_unifcube(n,\ p)} function generates \(n\) points uniformly distributed within a \(3\text{-}D\) cube centered at the origin, optionally embedded in a higher-dimensional space with additional noise dimensions.

Points are sampled independently from \(X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)\) of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in \(3\text{-}D\) space.

For \(p > 3\), the remaining dimensions are filled with Gaussian noise:

\[
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\]

\begin{verbatim}
unifcube <- gen_unifcube(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-unifcube-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `unifcube` data.}\label{fig:fig-unifcube-proj}
\end{figure}

The \texttt{gen\_cubehole(n,\ p)} function generates a dataset of \(n\) points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in \(p\)-dimensional space.

The initial points are drawn from a uniform cube centered at the origin: \(X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)\).

The center of the cube is \((0, 0, 0)\). A spherical hole of radius \(0.5\) is carved out by filtering out points whose Euclidean distance from the origin is less than or equal to \(0.5\):

Remove if \(\sqrt{X\_1^2 + X\_2^2 + X\_3^2} \leq 0.5\).

This results in a shell-like point cloud that fills the cube but avoids the spherical center.

For \(p > 3\), Gaussian noise is added to the remaining dimensions (via \texttt{gen\_unifcube()}):

\[
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\].

\begin{verbatim}
cubehole <- gen_cubehole(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubehole-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `cubehole` data.}\label{fig:fig-cubehole-proj}
\end{figure}

@@

The \texttt{gen\_gaussian(n,\ p,\ s)} function generates a multivariate Gaussian cloud in \(p\)-dimensional space, centered at the origin with user-defined covariance structure.

Each point is independently drawn using the multivariate normal distribution with:

\[
X_i \sim \mathcal{N}_p(\boldsymbol{0}, \Sigma)
\]

where \(\Sigma\) is a user-defined \(p \times p\) positive-definite matrix.

\begin{verbatim}
gau <- gen_gaussian(n = 1000, p = 4, s = diag(4))
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-gau-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `gau` data.}\label{fig:fig-gau-proj}
\end{figure}

@@

The \texttt{gen\_longlinear(n,\ p)} function simulates a high-dimensional linear structure with a dominant linear trend and small additive noise.

Each variable \(X_i\) is created as:

\[
X_i = \text{scale}_i \cdot \left(0, 1, \dots, n{-}1 + \epsilon\right) + \text{shift}_i
\]

where:

\(\text{scale}_i \sim \mathcal{U}(-10, 10)\): randomly chosen direction and stretch.

\(\text{shift}_i \sim \mathcal{U}(-300, 300)\): large offset to spread features apart.

\(\epsilon \sim \mathcal{N}(0, 0.03 \cdot n)\): Gaussian noise scaled to dataset size to introduce mild irregularity.

\begin{verbatim}
linear <- gen_longlinear(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-linear-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `linear` data.}\label{fig:fig-linear-proj}
\end{figure}

@@

The \texttt{gen\_mobius(n,\ p)} function generates a dataset of \(n\) points that form a Mobius strip embedded in the first three dimensions of a \(p\)-dimensional space. This classical non-orientable surface loops back on itself with a half-twist.

The Mobius strip is generated using \texttt{geozoo::mobius(n,\ p\ =\ 3)\$points}, which samples \(n\) points from a parametrization of the surface defined by angle \(\theta\) and strip width \(w\). The cartesian coordinates are;

\[
X\_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta),
\]
\[
X\_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta),
\]
\[
X\_3 = \frac{w}{2} \sin(\theta/2).
\]

This maps a \(2\text{-}D\) band with a half-twist into \(3\text{-}D\) space, forming a non-orientable 1-sided surface.

For \(p > 3\), the function appends \(p - 3\) dimensions with small Gaussian noise:

\[
x_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\]

\begin{verbatim}
mobius <- gen_mobius(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-mobius-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `mobius` data.}\label{fig:fig-mobius-proj}
\end{figure}

@@

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions of a \(p\)-dimensional space. This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let \(X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])\) be the independent variable. A raw polynomial basis of degree 2 is used to form \(X\_2 = X\_1 - X\_1^2 + \varepsilon\_2\), where \(\varepsilon\_2 \sim \mathcal{U}(0, 0.5)\).

This creates a smooth parabolic arc that opens downward, with jitter in the vertical direction to simulate measurement noise or intrinsic variability.

For \(p > 2\), the dataset includes Gaussian noise in the remaining dimensions via \texttt{gen\_noisedims()}:

\[
X\_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
\]

\begin{verbatim}
quadratic <- gen_quadratic(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-quadratic-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `quadratic` data.}\label{fig:fig-quadratic-proj}
\end{figure}

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions of a \(p\)-dimensional space. This function creates a more complex curvilinear structure than a simple parabola.

The shape is generated using a third-degree raw polynomial basis expansion. Let \(X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])\) be the base input. A cubic transformation with vertical jitter is used to define \(X\_2 = X\_1 + X\_1^2 - X\_1^3 + \varepsilon\_2\$, where \$\varepsilon\_2 \sim \mathcal{U}(0, 0.5)\).

When \(p > 2\), the remaining dimensions consist of independent Gaussian noise features generated via \texttt{gen\_noisedims()}:

\[
X\_j \sim \mathcal{N}(0, 0.1^2)\$ for \$j = 3, \dots, p.
\]

\begin{verbatim}
cubic <- gen_cubic(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cubic-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `cubic` data.}\label{fig:fig-cubic-proj}
\end{figure}

@@

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a \(p\)-dimensional space, with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l\_x, l\_y)\) to a smaller rectangular section with side lengths \((2r\_t, 2r\_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final dimension \(x\_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\):

\[
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)
\].

Let \(r\_x(z)\) and \(r\_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\):

\[
r_x(z) = r_t + (l_x - r_t) \cdot \frac{z}{h}, \quad
r_y(z) = r_t + (l_y - r_t) \cdot \frac{z}{h}
\].

The first three coordinates are then defined as:

\[
x_1 \sim \mathcal{U}(-r_x(z),\ r_x(z)), \quad
x_2 \sim \mathcal{U}(-r_y(z),\ r_y(z)), \quad
x_3 \sim \mathcal{U}(-r_x(z),\ r_x(z))
\].

For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(x\_4\) to \(x\_{p-1}\)) are tapered toward zero with decreasing height and generated as:

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
\]

This tapering ensures that the structure narrows in all directions as it approaches the tip.

\begin{verbatim}
pyrrect <- gen_pyrrect(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrrect-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrrect` data.}\label{fig:fig-pyrrect-proj}
\end{figure}

The \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a \(p\)-dimensional space, with the tip located at height 0 and the base at height \(h\). The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \$x\_p\$ encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

\[
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right).
\]

Let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\):

\[
r(z) = r_t + (l - r_t) \cdot \frac{z}{h}.
\]

A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section:

\[
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
\]

The first three coordinates (triangle plane) are then:

\[
x_1 = r(z)(1 - u - v), \quad
x_2 = r(z) \cdot u, \quad
x_3 = r(z) \cdot v
\]

For dimensions \(j = 4, \dots, p-1\), values are tapered linearly toward zero near the tip:

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
\]

The \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a \(p\)-dimensional space, with the tip located at height 0 and the base at height \(h\). The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \(x\_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

\[
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right).
\]

Let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\):

\[
r(z) = r_t + (l - r_t) \cdot \frac{z}{h}.
\]

A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section:

\[
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
\]

The first three coordinates (triangle plane) are then:

\[
x_1 = r(z)(1 - u - v), \quad
x_2 = r(z) \cdot u, \quad
x_3 = r(z) \cdot v
\]

For dimensions \(j = 4, \dots, p-1\), values are tapered linearly toward zero near the tip:

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
\]

The result is a tapered, high-dimensional triangle-based pyramid with more points near the apex and greater width at the base.

\begin{verbatim}
pyrtri <- gen_pyrtri(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrtri-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrtri` data.}\label{fig:fig-pyrtri-proj}
\end{figure}

The \texttt{gen\_pyrstar(n,\ p,\ h,\ rb)} function generates a dataset of \$n\$ points forming a high-dimensional pyramid-shaped structure with a hexagonal base. The pyramid extends vertically from height \(z = 0\) (tip) to height \(z = h\) (base), embedded in a \(p\)-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \(x\_p\) encodes the height \(z\):

\[
x_p = z \sim \mathcal{U}(0, h).
\]

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r\_b\):

\[
r(z) = r_b \cdot \left(1 - \frac{z}{h}\right).
\]

Each point is placed within a regular hexagon in the plane \((x\_1, x\_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}\) and a uniformly random radial scaling factor:

\[
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{\mathcal{U}(0, 1)}.
\]

Then, the first two coordinates are:

\[
x_1 = r(z) \cdot r_{\text{point}} \cdot \cos(\theta), \quad
x_2 = r(z) \cdot r_{\text{point}} \cdot \sin(\theta)
\]

For dimensions \(j = 3, \dots, p-1\), values are tapered toward zero near the tip:

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
\]

This results in a smooth, star-like pyramid in \(p\)-dimensional space, with a broad hexagonal base and narrow tip.

\begin{verbatim}
pyrstar <- gen_pyrstar(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrstar-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrstar` data.}\label{fig:fig-pyrstar-proj}
\end{figure}

The \texttt{gen\_pyrholes(n,\ p)} function generates \(n\) points embedded in a \(p\)-dimensional simplex using a chaotic attractor-like midpoint algorithm. The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a ``Sierpinski-like pyramid'' in high dimensions.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The generation process begins with an initial point \(T\_0 \in [0, 1]^p\) drawn from a uniform distribution:

\[
T_0 \sim \mathcal{U}(0, 1)^p.
\]

Let \(C\_1, C\_2, \dots, C\_{p+1}\) denote the corner vertices of a \(p\)-dimensional simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T\_{i-1}\) and a randomly selected vertex \(C\_k\):

\[
T_i = \frac{1}{2}(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}.
\]

This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex.

The final output is a \(n \times p\) matrix where each row represents a point:

\[
X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}.
\]

\begin{verbatim}
pyrholes <- gen_pyrholes(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-pyrholes-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `pyrholes` data.}\label{fig:fig-pyrholes-proj}
\end{figure}

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \texttt{n} observations in \texttt{p} dimensions.

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure:

\[
  X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1).
\]

This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction.

For \(p > 3\), additional noise dimensions are appended introducing structured, wavy perturbations.

\begin{verbatim}
scurve <- gen_scurve(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-scurve-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `scurve` data.}\label{fig:fig-scurve-proj}
\end{figure}

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)}. This function produces \texttt{n} observations in a \(p\)-dimensional space where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point \((0, 1, 0, \ldots)\) is defined in \(p\)-dimensional space. All observations within a Euclidean distance of \(\sqrt{0.3} \approx 0.5477\) from the anchor are removed:

\[
\text{remove if} \quad \sum_{j=1}^p (X_j - a_j)^2 \leq 0.3
\].

This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

\begin{verbatim}
scurvehole <- gen_scurvehole(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-scurvehole-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `scurvehole` data.}\label{fig:fig-scurvehole-proj}
\end{figure}

@@

The function \texttt{gen\_circle(n,\ p)} generates a \(p\)-dimensional dataset of \(n\) observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale.

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane:

\[
X_1 = \cos(\theta), \quad X_2 = \sin(\theta)
\]

For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\).

Each additional dimension is computed as:

\[
X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p
\]

This structure retains a one-dimensional manifold (the circle) but allows it to be embedded non-linearly and decorrelatively in higher dimensions. The decreasing scale ensures that higher dimensions add less variation, preserving the dominance of the circular shape while adding complexity.

\begin{verbatim}
circle <- gen_circle(n = 1000, p = 4)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-circle-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the `circle` data.}\label{fig:fig-circle-proj}
\end{figure}

The \texttt{gen\_curvycycle(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations. The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle.'' In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = \frac{1}{3} \cdot \cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.
